---
title: "NYDP Shooting Incident Data"
author: "Garey Salinas"
date: "2024-11-05"
output:
  html_document:
    toc: true           # Enables the Table of Contents
    toc_depth: 2        # Sets the depth to include up to ## headers (1 for #, 2 for ##)
    toc_float: true     # Makes the TOC float on the side (only for HTML output)
  pdf_document:
    toc: true           # TOC for PDF output as well
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## NY Shooting Incident Data

The NY Shooting Incident data set provides a comprehensive record of every shooting incident reported in New York City from 2006 through the end of the previous calendar year. The New York city data set is a csv file and can be downloaded from [https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD](https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD)

## Import Libraries

```{r import-libraries}
library(stringr)
library(readr)
library(tidyverse)
library(knitr)
```

## Load Data

I will start by reading in the data from the link provided above. 

```{r read_data, echo = TRUE}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
incidents <- read_csv(url, show_col_types = FALSE)
```


### Data

View data set

```{r view-incident-data}
incidents
```

Summary of data set

```{r summary-incident-data}
summary(incidents)
```

After looking at the data set, I want to tidy the data set by removing the `INCIDENT_KEY, X_COORD_CD, Y_COORD_CD, PRECINCT, LOC_OF_OCCUR_DESC, JURISDICTION_CODE, LOC_CLASSFCTN_DESC, LOCATION_DESC, Latitude, Longitude, Lon_Lat`.

```{r remove-cols}
incidents_clean <- incidents %>% select(-c(INCIDENT_KEY, 
                                           X_COORD_CD, 
                                           Y_COORD_CD, 
                                           PRECINCT, 
                                           LOC_OF_OCCUR_DESC, 
                                           JURISDICTION_CODE, 
                                           LOC_CLASSFCTN_DESC, 
                                           LOCATION_DESC, 
                                           Latitude, Longitude, Lon_Lat))
incidents_clean
```

Count the number of missing values in each column

```{r missing-values}
# Count NA values for each column in incidents_clean
na_counts <- sapply(incidents_clean, function(x) sum(is.na(x)))

# Convert the result to a DataFrame with specified column names
na_summary <- tibble(
  Columns = names(na_counts),
  NA_Count = na_counts
)

kable(na_summary)
```


Converting `OCCUR_DATE` object into a date object

```{r convert-occur-date}
incidents_clean$OCCUR_DATE <- mdy(incidents_clean$OCCUR_DATE)
incidents_clean
```

## Plots

Bar chart of shooting incidents by `BORO`

```{r boro-bar-chart}
ggplot(incidents_clean, aes(x = BORO, fill = BORO)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Count of Each BORO", x = "BORO", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")

```

Stacked bar chart of `STATISTICAL_MURDER_FLAG` (TRUE/FALSE) in each `BORO`

```{r barchart-murder-boro}
ggplot(incidents_clean, aes(x = BORO, fill = STATISTICAL_MURDER_FLAG)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)),
            position = position_stack(vjust = 0.5)) +
  labs(title = "Murder Counts by Boro", x = "Boro", y = "Count", fill = "MURDER") +
  theme_minimal()
```

Showing a line plot of Brooklyn Murders.

```{r brooklyn-murders-yearly}
# Filter for Brooklyn murders and aggregate by year
brooklyn_murders_yearly <- incidents_clean %>%
  filter(BORO == "BROOKLYN" & STATISTICAL_MURDER_FLAG == TRUE) %>%  # Filter for Brooklyn murder incidents
  mutate(year = year(OCCUR_DATE)) %>%                                # Extract year from date
  group_by(year) %>%                                                 # Group by year
  summarize(total_incidents = n())                                   # Count murders per year

# Plot the line chart
ggplot(brooklyn_murders_yearly, aes(x = year, y = total_incidents)) +
  geom_line(color = "red") +                         # Line plot for yearly totals
  geom_point(color = "black", size = 2) +            # Add points at each year for clarity
  labs(title = "Yearly Murder Totals in Brooklyn",
       x = "Year",
       y = "Total Murders") +
  theme_minimal()
```

## Analysis
 - Brooklyn has the highest total incidents among all the boroughs.
 - The Bronx and Brooklyn have the highest murder counts, with 1,634 and 2,210 murders. 
 - Manhattan and Queens have moderate murder counts, while Staten Island has the lowest murder count. 
 - The proportion of murders to non-murders varies between boroughs. For example, while Brooklyn has the highest number of murders, it also has a very high count of non-murders. 
 - Murder incidents appear to be highly variable across the years. 
 - The later years in the data set, especially 2020 onward, show relatively lower and more consistent incident counts.
 -  Incidents peaked from 2008 to 2010, with a decrease in incidents from 2011 to 2019,  then spiking again around 2020.
 

## Bias 
- The analysis did not account for socioeconomic and demographic factors. Income, employment rates, and population density can influence incident levels. 
- Some neighborhoods may experience higher police presence and higher reporting rates, which can skew the data toward these areas.  
- Not all crimes could have been reported, especially in under-resourced communities.  

## Analysis Conclusion
The Project examined the murder trend in Brooklyn. My analysis identified several key findings. Brooklyn has the highest number of murders; it also has a very high count of non-murders. Incidents peaked from 2008 to 2010, with decreased incidents from 2011 to 2019,  then spiking again around 2020. High murder rates could indicate a period of economic depression, such as the mortgage crisis and the COVID-19 pandemic. Other factors, such as employment and other socioeconomic factors, contribute to the number of Incidents. Periods of low or no incidents suggest effective policing or community engagement.  

## Question
- Can we provide a model that predicts the number of murder incidents in Brooklyn to help law enforcement target resources more effectively?

## Model
The Brooklyn yearly trend seem to be non-linear so we will model the yearly trends using a polynomial regression of degree 2 or degree 3 and check which one is a better fit. 

```{r murder-model}
# Fit a polynomial regression model of degree 2
model_poly2 <- lm(total_incidents ~ poly(year, 2), data = brooklyn_murders_yearly)

# Fit a polynomial regression model of degree 3
model_poly3 <- lm(total_incidents ~ poly(year, 3), data = brooklyn_murders_yearly)

# Original line plot
p <- ggplot(brooklyn_murders_yearly, aes(x = year, y = total_incidents)) +
  geom_line(color = "red") +
  geom_point(color = "black", size = 2) +
  labs(title = "Yearly Murder Totals in Brooklyn with Regression Model",
       x = "Year",
       y = "Total Murders") +
  theme_minimal()

# Add the polynomial regression line (degree 2)
p + geom_smooth(method = "lm", formula = y ~ poly(x, 2), color = "blue", se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), color = "green", se = FALSE)
```
 
## Model Summary
```{r model-summary}
# Summary of the quadratic model
summary(model_poly2)

# Summary of the cubic model
summary(model_poly3)

```

Based on the similar R-squared, higher RSE, and non-significant cubic term, the quadratic model (degree 2) is a better choice. It provides a similar fit with fewer terms, making it more straightforward and interpretable.


Using AIC and BIC to help choose the best model
```{r choose-model}
AIC(model_poly2, model_poly3)
BIC(model_poly2, model_poly3)
```

The quadratic model is better based on lower AIC and BIC values, simplicity, and interpretability. 

## Prediction
```{r brooklyn-murder-prediction}
# Create a data frame for the year 2025
murder_prediction <- data.frame(year = 2025)

predicted_value_2025 <- predict(model_poly2, newdata = murder_prediction)

print(predicted_value_2025)
```

Model predicts that 121 murders will occur in Brooklyn in the year 2025. 

## Session Information

```{r session-info}
sessionInfo()
```